A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
dropped some needless 'will's

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types, even if they occur rarely in practice. A type
may also be a corecursive singleton only in some models. For example, if
\ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be singleton types
precisely for the models that interpret \ty{unit} as a singleton. Fortunately,
given cardinalities for the ordinary types, it is easy to characterize
this degenerate case:

@@

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms.
We will write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

We present the calculus in three steps.
In the first step, we compute the bidirectional closure of $\Ec$;
in the second step, we make inferences based on cyclicity and bisimilarity;
and in the third step, %(when necessary),
we branch on constructor types for various terms in $\tEc$.
Following the conventions from~\cite{}, 
the derivation rules of our calculus are given in \emph{guarded assignment form},
where a rule can be applied to $\Ec$ if it meets all of the specified preconditions for $\Ec$.
The conclusion of a rule either describes equalities to be added to $\Ec$ (in which case we will call an application of it to be \emph{nonterminal}),
or is $\bot$ (in which case we will call an application of it to be \emph{terminal}).
A rule may have multiple conclusions separated by $\parallel$, which denotes nondeterministic branching.

@@

\paragraph{Step 2: Applying Cyclicity and Bisimilarity.}
In the second step,
we assume that $\Ec$ is saturated with respect to the rules in Figure~\ref{fig:cc-rules}---that is,
no rule can be applied to extend $\Ec$ further.
When this is the case, it is easy to see that $\Ec$ induces an equivalence relation over $\tEc$ such that two terms $t_1$ and $t_2$ are equivalent if and only if $t_1 \teq t_2 \in \Ec$.
Thus, we will in the following consider $\Ec$ as a set of equivalence classes of terms. 
For term $t \in \tEc$, we will write $\ec{t}$ to denote the equivalence class in $\Ec$ that contains $t$.

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
For instance, 
$\mu x. \const{C}( x ) \vsim \mu y. \const{C}( y )$ and
$\mu x, \const{D}( y, x ) \vsim \mu z. \const{D}( y, z )$, but
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{C}( x )$,
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{D}( z, x )$ and
$\mu x. \const{D}( y, x ) \not\vsim \mu y. \const{D}( y, x )$.
Also, note that
two $\mu$-terms may represent the same value while being not bisimilar, as in
$\mu x. \const{C}( \mu y. \const{C}( x ) ) \not\vsim \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. \const{C}( \vec u ) \in t_1$ indicates that $\mu x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. \const{C}( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & \mu x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.

@@

For any $\mu$-term $t$ of the form $\mu x. C( \vec u )$, we may recursively construct a normal $\mu$-term $\nf{t}$
(which we will call the \emph{normal form} of $t$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$,
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we will write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

>>>>

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types, even if they rarely arise in practice. A type
may also be a corecursive singleton only in some models. For example, if
\ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be singleton types
precisely for the models that interpret \ty{unit} as a singleton. Fortunately,
given cardinalities for the ordinary types, it is easy to characterize
this degenerate case:

@@

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

We present the calculus in three steps.
In the first step, we compute the bidirectional closure of $\Ec$;
in the second step, we make inferences based on cyclicity and bisimilarity;
and in the third step, %(when necessary),
we branch on constructor types for various terms in $\tEc$.
Following the conventions from~\cite{}, 
the derivation rules of our calculus are given in \emph{guarded assignment form},
where a rule can be applied to $\Ec$ if it meets all of the specified preconditions for $\Ec$.
The conclusion of a rule either describes equalities to be added to $\Ec$ (in which case we call an application of it \emph{nonterminal}),
or is $\bot$ (in which case we call an application of it \emph{terminal}).
A rule may have multiple conclusions separated by $\parallel$, which denotes nondeterministic branching.

@@

\paragraph{Step 2: Applying Cyclicity and Bisimilarity.}
In the second step,
we assume that $\Ec$ is saturated with respect to the rules in Figure~\ref{fig:cc-rules}---that is,
no rule can be applied to extend $\Ec$ further.
When this is the case, it is easy to see that $\Ec$ induces an equivalence relation over $\tEc$ such that two terms $t_1$ and $t_2$ are equivalent if and only if $t_1 \teq t_2 \in \Ec$.
Thus, we can consider $\Ec$ as a set of equivalence classes of terms. 
For a term $t \in \tEc$, we write $\ec{t}$ to denote the equivalence class in $\Ec$ that contains $t$.

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure as either a variable $x$
or a node of the form
$\mu x.\; \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$.
For uniformity, we also consider $\mu x.\; \const{c}$ to be a $\mu$-term
if $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are well-typed---i.e., whose function symbols
are applied to arguments of the expected type.
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables.
A $\mu$-term is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables,
in which case $t$ and $s$ are $\alpha$-\emph{equivalent}.
For instance, 
$\mu x. \const{C}( x ) \vsim \mu y. \const{C}( y )$ and
$\mu x, \const{D}( y, x ) \vsim \mu z. \const{D}( y, z )$, but
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{C}( x )$,
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{D}( z, x )$ and
$\mu x. \const{D}( y, x ) \not\vsim \mu y. \const{D}( y, x )$.
Also, note that
two $\mu$-terms may represent the same value while being not bisimilar, as in
$\mu x. \const{C}( \mu y. \const{C}( x ) ) \not\vsim \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. \const{C}( \vec u ) \in t_1$ indicates that $\mu x. \const{C}( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. \const{C}( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. \const{C}( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & \mu x. \const{C}( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. \const{C}( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.

@@

For any $\mu$-term $t$ of the form $\mu x. C( \vec u )$, we may recursively construct a normal $\mu$-term $\nf{t}$
(which we call the \emph{normal form} of $t$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$,
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
We compute the \emph{completion} of $\Val$ for normal model $\M$, which we write as $\Val^\ast$,
which we construct by assigning values from $\M$ to unassigned variables in the domain of $\Val$.
